---
title: "IPM_champ"
author: "ZZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: no
      smooth_scroll: no
---

First come, first served project\\

sp: 1 = *Chlamy*, 2 = *Acutodesmus obliquus*, 3 = *Scenedesmus quadricauda*, 4= *Monoraphidium griffithii*, 5 = *Chlorella vulgaris*

```{r setup}
knitr::opts_chunk$set(message = F,warning = F)
```




```{r}
options(tidyverse.quiet = TRUE)
library(tidyverse)
library(keras)
library(vegan)
library(lmerTest)
dir.create('fig') # create
source('functions.R')

# theme for ggplot
theme <- theme_bw() + 
  theme(panel.background = element_blank(),
              panel.border     = element_rect(fill = NA), 
              strip.background = element_blank(),
              strip.text.x = element_text(size = 8),
              axis.text.x = element_text(size = 8, angle = 0, hjust = 0.5, colour = 'black'),
              axis.text.y  = element_text(size = 8, colour = 'black'),
              axis.title = element_text(size = 9),# size of x  y lab
              #legend.position   = c(0.9, 0.85),
              legend.title      = element_text(size = 8),
              legend.text       = element_text(size = 6),
              legend.key.height = unit(0.8,'line'),
              legend.background = element_blank(),
              legend.key        = element_rect(colour = NA, fill = NA),
              plot.tag    = element_text(size = 9.5, face  = 'bold'),
              plot.tag.position = c(0.03, 1),
              plot.title  = element_text(hjust = 0.5, size = 10),
              plot.margin = margin(t = 0.5, r = 0.1, b = 0, l = 0.1, unit = "cm")) 

```


# read and clean data
```{r, eval = F}
# read the competition assays
file_names <- list.files('dat/')

my_read_delim <- function(file, path = 'dat') { # function to read the each plate
  d <- read.delim(file.path(path, file), header = T, skip = 5)
  return(d)

}

dat_raw <- (lapply(file_names, function(file) my_read_delim(file))) %>%
  bind_rows

rm(my_read_delim)

save(dat_raw, file = 'd_raw20250326.RData')# add 24 new files from t11
```

check whether there are exporting problems (e.g. some of the wells not exported )
```{r, eval = F}
#load('d_raw.RData')
load('d_raw.RData')
dat_raw %>% select(MEASUREMENT.SET.ID, Well.Name) %>% distinct() %>% group_by(MEASUREMENT.SET.ID) %>% tally() 
# the well that does not have culture
dat_raw %>% filter(MEASUREMENT.SET.ID %in% c('33421', '33422')) %>% select(MEASUREMENT.SET.ID, Well.Name) %>% distinct() %>% 
  group_by(Well.Name) %>% tally %>% filter(n == 1) %>% 
  pull(Well.Name)
```

info\
t0 and t_n are different, need to clean separately\
t_n
```{r, eval = F}
info_medium   <- tibble(medium  = c(3, 2, 1, 2, 0, 0, 3, 1), # match column with medium
                        column = paste(0, 1:8, sep ='')
                      ) %>% 
  # the detail info of the medium
  left_join(tibble(medium = 0:3,
                   no     = c(64, 0, 48, 16),
                   nh     = c(0, 64, 16, 48))
            )


# the plate order on day 1 and day 4
info_plate <- tibble(MEASUREMENT.SET.ID = dat_raw$MEASUREMENT.SET.ID %>% unique %>% sort , # plate_ID from IXM
                     culture = c(NA,  # t0 plate has both mono and community, so I assigned IT NA
                                 c('mono', rep('comp', 5)) %>% rep(3), # t1 - t3
                                 rep('comp', 5), 'mono', # t4
                                 c('mono', rep('comp', 5)) %>% rep(6 + 4)  # t5 again, first plate is the monocutlure
                                 ), # t0 plate has both mono and community, so I assigned IT NA
                     comm_v  = c(NA, 
                                 0:5 %>% rep(3), #t1-t3
                                 1:5, 0, # t4
                                 0:5 %>% rep(6 + 4)
                                 ), # e.g. 1 means combination 1-8
                     day = c(0, rep(1:14 * 3.5,  each = 6))
                       ) 

# monoculture plate, match species with the row
info_mono <- tibble(sp = 1:5,
                    row = LETTERS[1:5]
                    )

# somehow the R code did not give the same random number, So, i have to copy from the python code
info_comm <- tibble(comm = c(23, 4, 19, 37, 34, 3, 6, 27, 29, 18, 5, 21, 32, 39, 
                        14, 30, 26, 17, 35, 33, 15, 28, 0, 22, 23, 4, 19, 
                        7, 34, 3, 6, 13, 29, 18, 5, 8, 32, 39, 14, 36, 26, 20, 16, 
                        2, 15, 24, 31, 9, 23, 4, 37, 7, 34, 3, 27, 13, 29, 18, 21, 8, 
                        32, 39, 30, 36, 17, 20, 11, 25, 28, 24, 38, 1, 23, 19, 37, 7, 34,
                        6, 27, 13, 29, 5, 21, 8, 32, 14, 30, 36, 35, 16, 11, 10, 0, 31, 38,
                        12, 4, 19, 37, 7, 3, 6, 27, 13, 18, 5, 21, 8, 39, 14, 30, 36, 33, 2, 25, 10, 22, 9, 1, 12),
               
               sp_ID = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4)
               ,
               vol = c(165, 165, 165, 165, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 198, 198, 198, 198, 22, 22, 22, 22, 18.3333333333333, 18.3333333333333, 18.3333333333333, 165, 165, 165, 165, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 22, 198, 198, 198, 198, 22, 22, 22, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 165, 165, 165, 165, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 22, 22, 198, 198, 198, 198, 22, 22, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 165, 165, 165, 165, 18.3333333333333, 18.3333333333333, 18.3333333333333, 22, 22, 22, 198, 198, 198, 198, 22, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 18.3333333333333, 165, 165, 165, 165, 22, 22, 22, 22, 198, 198, 198, 198)
) %>% 
  mutate(comm = comm + 1, # python to R language
         sp_ID = sp_ID + 1) %>% 
  rename(sp = sp_ID) %>% 
  arrange(comm) %>% 
  group_by(comm) %>% 
  mutate(high = ifelse(vol == max(vol), sp, 0)) %>%  # ‘high’ tells with species has the highest ratio
  summarise(lst_sp = paste(sp, collapse = '_'),
            n_sp   = n(),
            high = max(high)) %>% 
  mutate(comm_v = rep(1:5, each  = 8),
         row    = rep(LETTERS[1:8], 5))
```


t0
```{r, eval = F}
info_mono_t0 <- tibble(sp = 1:5, 
                       column = c('07', '08', '09', '10', '11')
                       )

info_comm_t0 <- info_comm %>% select(-comm_v, -row) %>% 
  mutate(Well.Name = outer(LETTERS[1:8], 1:5, paste, sep = '0') %>% as.vector())
```


```{r, eval = F}
dat_cl <- dat_raw %>% 
  mutate(column = str_sub(Well.Name, 2, 3),
         row    = str_sub(Well.Name, 1, 1)
         ) %>% 
  left_join(info_plate)

```





```{r, eval = F}
# monoculture for training
d_mono <- bind_rows(dat_cl %>% filter(day > 0, culture == 'mono') %>% left_join(info_mono), # t_n
                    dat_cl %>% filter(day == 0) %>% right_join(info_mono_t0) # t0
                    )
# distinguish the species 
d_comp <- bind_rows(dat_cl %>% filter(day > 0, culture == 'comp') %>% 
                      left_join(info_comm) %>% left_join(info_medium),
                    dat_cl %>% filter(day == 0) %>% right_join(info_comm_t0)
                    ) %>% 
  mutate(comm_v = ifelse(is.na(comm_v), 0, comm_v)) %>% 
  filter(!(comm_v == 1 & Well.Name == 'B02'))

```




use only necessary data (e.g. features, day, which species, which ratio, which medium)
```{r, eval = F}
list <- colnames(d_comp)[str_detect(colnames(d_comp), 'Cell\\.\\.')]
list <- list[!list %>% str_detect('ObjectID|Features.Count_Sum|Line.Length_Average')]



d_comp <- d_comp %>% 
  dplyr::select(all_of(list),  
                Well.Name, comm, day, no, nh, high, lst_sp, n_sp, medium ) # well name is needed, as this tell the replicate
d_mono  <- d_mono %>% 
  dplyr::select(all_of(list),  
                Well.Name, sp) 
save(d_comp, d_mono, list, theme, file = 'd_cl20250327.RData')
```


# ML to distinguish differnt sp




```{r, eval = F}
# load('d_cl.RData')
load('d_cl20250327.RData')
lst_comm <- d_comp %>% pull(lst_sp) %>% unique
```



```{r, eval = F}
lst_res <- list()
pb = txtProgressBar(min = 0, max = length(lst_comm), initial = 0, style = 3) 
for (comm_i in lst_comm) {
  res_i <- deep_lear_n_sp(community = comm_i)
  lst_res <- c(lst_res, res_i)
  setTxtProgressBar(pb, which(comm_i == lst_comm))
}
save(lst_res, lst_comm, file = 'd_comp_obs20250327.RData') 
```




# data exploration
```{r}
load('d_comp_obs20250327.RData')
```


```{r}
# accuracy of deep learning models
t_acc_dl <- tibble(comm = lst_comm[1:(length(lst_res)/3)],
                   acc = lst_res[seq_along(lst_res) %% 3 == 1] %>% unlist
                )
print(paste('median accuracy:', t_acc_dl$acc %>% median))

print('low accuracy combinations')
t_acc_dl %>% arrange(acc) %>% head(4) # it is difficult to distinguish between sp 2 and 3
```


```{r}
d_comp_cl <- lst_res[seq_along(lst_res) %% 3 == 0] %>% # only take the third sub-list, which is the data
  bind_rows()
# save(d_comp_cl, file = 'champ_abun.RData')
```

## day 0 
```{r}
d_comp_cl %>% filter(day == 0) %>% 
  group_by(lst_sp, high, n_sp) %>% 
  mutate(freq = n/sum(n)) %>% 
  ggplot(aes(sp, freq)) +
  geom_point() + 
  facet_grid(n_sp + lst_sp~high) + 
  theme
# ggsave(file = 'fig/freq_ini.pdf', height = 25/2.54, width = 15/2.54)
```

## how the frequency of the initial abundant species change over time
```{r, fig.height= 20/2.54, fig.width= 15/2.54}
d_comp_cl %>% 
  filter(n_sp == 2) %>% 
  filter(medium != 'NA') %>%
  group_by(lst_sp, high, n_sp, medium, Well.Name, day) %>% 
  mutate(freq = n/sum(n)) %>% 
  filter(sp == high) %>% 
  ggplot(aes(day, freq, color = medium %>% as.factor())) +
  geom_point(size = 0.5, alpha = 0.5) + 
  facet_grid(lst_sp ~ high) + 
  theme +
  theme(legend.position = 'top')

ggsave(file = 'fig/freq.pdf', height = 25/2.54, width = 15/2.54)
```

pivot wider the abundance data
```{r}
lst_comp <- d_comp_cl %>% 
  mutate(col = str_sub(Well.Name, 2, 3)) %>% 
  pivot_wider(names_from = sp, values_from = n, values_fill = 0) %>% 
  arrange(lst_sp, day, col, no, high) %>% 
  group_split(lst_sp, no, nh, medium, day, col
              )
```





# analyses

```{r}
cal_beta <- function(data){
  if(nrow(data) == 1){
    data$sim <- NA
  }
  
  if(nrow(data) != 1){
        
    sim <- 1 -data %>% select(`1`, `2`, `3`, `4`, `5`) %>% vegdist(upper = T)# beta similarity between the communities of different initial conditions
    sim <- sim %>% as.matrix() 
    diag(sim) <- NA
    
    data$sim <- rowMeans(sim, na.rm = T) # order is already aligned before
    
  }

  return(data)
}

d_sim <- lapply(lst_comp, cal_beta)
d_sim <- d_sim %>% bind_rows()
d_sim <- d_sim %>% filter(!is.na(sim)) %>% 
  mutate(medium = paste('NO3: ', no, '; NH4: ', nh),
         n_res = ifelse(no *nh == 0, 1, 2),
         n_res = ifelse(is.na(n_res), 'initial', n_res)
         ) %>% 
  mutate(sp1 = ifelse(str_detect(lst_sp, '1'), 1, 0),
         sp2 = ifelse(str_detect(lst_sp, '2'), 1, 0),
         sp3 = ifelse(str_detect(lst_sp, '3'), 1, 0),
         sp4 = ifelse(str_detect(lst_sp, '4'), 1, 0),
         sp5 = ifelse(str_detect(lst_sp, '5'), 1, 0)
         )

```

## 2 sp
```{r}
d_sim2 <- d_sim %>% filter(n_sp == 2) %>% 
  select(n_sp, medium, no, nh, day, sim, lst_sp, col, n_res, sp1, sp2, sp3, sp4, sp5) %>% 
  distinct 
```


whether number of niches (resources) matters (only use the last day)
```{r}
m2 <- lmer(sim ~ n_res + (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
anova(m2)
```

whether presence of a certain species matters
```{r}
m2_sp1 <- lmer(sim ~ sp1 * n_res+ (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
m2_sp2 <- lmer(sim ~ sp2 * n_res+ (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
m2_sp3 <- lmer(sim ~ sp3 * n_res+ (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
m2_sp4 <- lmer(sim ~ sp4 * n_res+ (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
m2_sp5 <- lmer(sim ~ sp5 * n_res+ (1|lst_sp) + (1|medium), data = d_sim2 %>% filter(day == max(day)))
```


```{r}
anova(m2_sp1)
anova(m2_sp2)
anova(m2_sp3)
anova(m2_sp4)
anova(m2_sp5)
```

## 4 sp

```{r}
d_sim4 <- d_sim %>% filter(n_sp == 4, day == max(day)) %>% 
  select(n_sp, medium, no, nh, day, sim, lst_sp, col, n_res, high, sp1, sp2, sp3, sp4, sp5) %>% 
  distinct 
```

```{r}
m4 <- lmer(sim ~ n_res + (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))
anova(m4)
```


```{r}

m4_sp1 <- lmer(sim ~ sp1 * n_res+ (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))
m4_sp2 <- lmer(sim ~ sp2 * n_res+ (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))
m4_sp3 <- lmer(sim ~ sp3 * n_res+ (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))
m4_sp4 <- lmer(sim ~ sp4 * n_res+ (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))
m4_sp5 <- lmer(sim ~ sp5 * n_res+ (1|lst_sp) + (1|medium), data = d_sim4 %>% filter(day == max(day)))


anova(m4_sp1)
anova(m4_sp2)
anova(m4_sp3)
anova(m4_sp4)
anova(m4_sp5)


```

## all together

```{r}
d_sim_cl <- bind_rows(d_sim2, d_sim4) %>% 
  mutate(n_sp = as.factor(n_sp),
         n_res = as.factor(n_res)
         )

m_sp1 <- lmer(sim ~ sp1 * n_res * n_sp + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))
m_sp2 <- lmer(sim ~ sp2 * n_res * n_sp + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))
m_sp3 <- lmer(sim ~ sp3 * n_res * n_sp + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))
m_sp4 <- lmer(sim ~ sp4 * n_res * n_sp + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))
m_sp5 <- lmer(sim ~ sp5 * n_res * n_sp + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))

anova(m_sp1)
anova(m_sp2)
anova(m_sp3)
anova(m_sp4)
anova(m_sp5)
```



```{r, fig.width= 10}

bind_rows(d_sim2, d_sim4) %>% 
  mutate(day = as.factor(day)) %>% 
  ggplot(aes(day , sim, color  = n_res %>% as.factor())) +
  geom_boxplot() +
  scale_color_manual(values=c("blue", "red", 'black')) +
  facet_wrap( ~ n_sp) + 
  theme




```



```{r}
m <- lmer((sim) ~ as.factor(n_res) * as.factor(n_sp) + (1|lst_sp) + (1|medium), data = d_sim_cl %>% filter(day == max(day)))


anova(m)
```


```{r}
d_sim_cl %>% filter(day == max(day)) %>% group_by(n_sp,n_res) %>% summarise(sim = median(sim))
```
